---
title: "Bivariate Copula Regression with gamboostLSS"
author: "Bjarne Herbst"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Bivariate Copula Regression with gamboostLSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
library(gamboostLSS)
```

## Introduction 
`gamboostLSS` fits one additive predictor per distribution parameter. With 
`CopulaFamilies` this is extended to a bivariate response: each margin keeps its
own parameters, and an additional parameter `theta` describes how the two 
responses depend on each other. Every one of these parameters gets its own 
predictor and its own variable selection. 

By Sklar's theorem the joint density of two continuous responses factorises into
a copula density evaluated at the marginal CDFs and the two marginal densities,

$$
f(y_1, y_2) = c(F_1(y_1), F_2(y_2);\theta)f_1(y_1)f_2(y_2)
$$

so the joint log-likelihood is a sum of a dependence term and two marginal terms.
This is what `CopulaFamilies()` boosts. 

## A first model
The response is a two-column matrix. Everything else works exactly as in a normal
`gamboostLSS()` call.

```{r}
set.seed(1)
n <- 150
x <- rnorm(n)
eps <- matrix(rnorm(2 * n), n, 2) %*% chol(matrix(c(1, 0.3, 0.3, 1), 2, 2))
y <- cbind(x + eps[, 1], 0.5 * x + eps[, 2])

cf <- CopulaFamilies(GaussianLSS(), GaussianLSS(), copula = "clayton")
fit <- gamboostLSS(y ~ bbs(x), families = cf, data = data.frame(x = x),
                   control = boost_control(mstop = 50, nu = 0.1))

names(fit)
```

Five sub-models: two parameters per margin and the copula parameter. The suffixes 
`1` and `2` indicate which margin a parameter belongs to. 

`predict()` with `type = "response"` returns every parameter on its natural 
scale, including `theta` on the copula's own scale. 

```{r}
pr <- predict(fit, newdata = data.frame(x = c(-1, 0, 1)), type = "response")
round(cbind(mu1 = pr$mu1, sigma1 = pr$sigma1, theta = pr$theta), 3)
```

## Choosing a copula 
Four copulas are available. They differ in how they distribute dependence across 
the range of the data, in particular the tails. 

```{r}
data.frame(
  copula = c("gaussian", "clayton", "gumbel", "frank"),
  range = c("(-1,1)", "(0, Inf)", "[1, Inf)", "R \\ {0}"),
  tail = c("none", "lower", "upper", "none")
)
```

Clayton captures joint *low* values, Gumbel joint *high* values. 
Gaussian and Frank are symmetric. Note that Clayton and Gumbel cannot represent
negative dependence at all. 

Because `theta` lives on a different scale in each family, the fitted values are 
not comparable across copulas. `copula_tau()` converts them to Kendall's tau, 
which is 

```{r}
taus <- sapply(c("clayton", "gumbel", "frank"), function(cop){
  cf <- CopulaFamilies(GaussianLSS(), GaussianLSS(), copula = cop)
  fit <- gamboostLSS(y ~ bbs(x), families = cf, data = data.frame(x = x),
                     control = boost_control(mstop = 50, nu = 0.1))
  c(theta = mean(predict(fit, type = "response")$theta),
    tau = mean(copula_tau(fit)))
})
round(taus, 3)
```

The `theta` row is not comparable across columns, while the `tau` row is.

## Discrete and binary margins 
The marginal combination is detected automatically, and the likelihood is built
accordingly - from copula CDF differences for discrete margins and from the 
`h`-function (conditional copula CDF) for mixed cases.

```{r}
set.seed(42)
y_mix <- cbind(rbinom(n, 1, plogis(x)), rnorm(n))
cf_mix <- CopulaFamilies(BernoulliLSS(), GaussianLSS(), copula = "frank")
fit_mix <- gamboostLSS(y_mix ~ bbs(x), families = cf_mix, 
                       data = data.frame(x = x),
                       control = boost_control(mstop = 30, nu = 0.1))
names(fit_mix)
```

The binary margin contributes a single parameter, so this model has four 
sub-models rather than five.

## Selecting `mstop`
`mstop` is the main tuning parameter. `make.grid()` builds the full factorial 
grid over all parameters, which is infeasible - five parameters at ten candidate
values each is $10^5 = 100000$ combinations. `make_copula_grid()` instead 
searches a single shared stopping iteration. 

```{r}
grid <- make_copula_grid(fit, from = 10, by = 10)
dim(grid)
cvr <- cvrisk(fit, folds = cv(model.weights(fit), type = "subsampling", B = 3),
              grid = grid, papply = lapply, trace = FALSE)
mstop(cvr)
```

The selected values is applied with:

```{r}
invisible(fit[mstop(cvr)])
mstop(fit)
```

Note that `[` modifies the model in place. 

## A worked example
The `india` dataset records childhood malnutrition in India (see `gamboostLSS_Tutorial` for more information). We model two child health outcomes jointly: `stunting` (height-for-age z-score) and `cbmi` (BMI of the child), as smooth functions of the child's age `cage`.

```{r}
data(india)
y_india <- cbind(india$stunting, india$cbmi)

cf_india <- CopulaFamilies(GaussianLSS(), GaussianLSS(), copula = "gaussian")
fit_india <- gamboostLSS(y_india ~ bbs(cage), families = cf_india, data = india,
                         control = boost_control(mstop = 100, nu = 0.1))

fv <- lapply(fit_india, fitted)
round(c(mu1 = mean(fv$mu1), mu2 = mean(fv$mu2)), 3)
round(c(observed1 = mean(india$stunting), observed2 = mean(india$cbmi)), 3)
```

The fitted means recover the observed ones. The interesting part is the dependence, which is allowed to vary with age: 

```{r}
tau_india <- copula_tau(fit_india)
plot(india$cage, tau_india, pch = 16, cex = 0.4, xlab = "child age (months)",
     ylab = expression(hat(tau)))
abline(h = cor(india$stunting, india$cbmi, method = "kendall"), col = "red", 
       lty = 2)
```

The dashed line is the unconditional Kendall's tau of the two responses. The fitted curve shows how the association changes with age - something a single correlation coefficient cannot express.

## Practical Notes
**Starting values for theta**
The offset for `theta` is derived from the empirical Kendall's tau of the responses, inverted through the copula's tau-to-theta relationship. When that value lies outside what a copula can represent (e.g. negative dependence with Clayton) a static default is used instead. 

**Numerical stability** 
Very strong dependence combined with a large step size can destabilise the fit. If a fit aborts with "Infinite residuals" or "could not fit base-learner", reduce `nu`! The same data usually fits at `nu = 0.05` or `0.01`. 

## Session info

```{r}
sessionInfo()
```
